---
title: "What the Colombian drinking-water quality file measures, and how much of it is there"
subtitle: "Answers to questions 1 and 2 on the *Calidad del Agua para Consumo Humano en Colombia* dataset"
author:
  - name: Adriana Clavijo Daza
    affiliations:
      - name: ETH Zurich
date: last-modified
abstract: |
  This note answers the first two questions posed for the Colombian drinking-water quality file held in this repository. Question 1 asks for the observational unit. It is the municipality-year: each row carries the annual Water Quality Risk Index (IRCA) of one municipality in one calendar year, with separate values for the whole municipality, its urban area and its rural area, and department-year aggregates are stacked in the same file. Question 2 asks how complete the data are. The overall IRCA is never missing and every calendar year from 2007 to 2024 is present, but the rural index is unavailable in about half of the municipal rows, about 6% of the municipality-years that the national municipal register implies are absent, two thirds of municipalities have an uninterrupted series, and Chocó is missing more than four fifths of its municipality-years.
keywords: [water quality, IRCA, Colombia, observational unit, missing data, panel data]
---

TODO: clean the comments, only leave those that explain a decision

```{r setup}
#| include: false
# _quarto.yml makes manuscript/ look like a project root to here(); pin the repo root.
here::i_am("manuscript/index.qmd")
source(here::here("R", "prepare.R"))
# Packages beyond those loaded by prepare.R: ggplot2, scales, knitr, sf, patchwork
# (DOCX and HTML); leaflet, reactable, htmltools (HTML only). There is no renv;
# install missing ones from CRAN, e.g. install.packages("leaflet").
library(ggplot2)
library(scales)
library(knitr)
library(sf)
library(patchwork)

# Short department labels for figures
dept_short <- dept_names |>
  transmute(dept_code,
            dept_label = case_when(dept_code == "88" ~ "San Andrés y Providencia",
                                   dept_code == "11" ~ "Bogotá D.C.",
                                   TRUE ~ dept_name))

# Palette: one blue ramp for magnitudes, fixed categorical hues for series
blue_ramp <- c("#cde2fb", "#86b6ef", "#3987e5", "#1c5cab", "#0d366b")
grey_none <- "#e1e0d9"
grey_na <- "#f4f3ef"
col_urban <- "#2a78d6"; col_rural <- "#eb6834"; col_both <- "#1baf7a"
theme_set(theme_minimal(base_size = 8) +
            theme(panel.grid.minor = element_blank(),
                  panel.grid.major = element_line(colour = "#e1e0d9", linewidth = 0.25),
                  axis.ticks = element_blank(),
                  legend.key.size = unit(0.35, "cm")))
```

# Data and validation

The repository holds one file, `Calidad_del_Agua_para_Consumo_Humano_en_Colombia_20260915.csv`, published on Colombia's open-data portal from the national drinking-water surveillance system [@sivicap2026]. Territorial health authorities sample water supplies through the year, and the samples are consolidated into the *Índice de Riesgo de la Calidad del Agua para Consumo Humano* (IRCA), a 0 to 100 score of deviation from the standards of Resolution 2115 of 2007, classed into five risk bands: no risk (0 to 5), low (5.1 to 14), medium (14.1 to 35), high (35.1 to 80) and sanitarily unviable (above 80) [@res2115].

The file was imported with every column as text, and the structure that the rest of this note relies on was asserted in code rather than assumed: the column set, the row count, the absence of empty cells, the format of the codes, the parsing of the year, the fact that the token `ND` is the only non-numeric value in the index columns, the spelling set of the risk labels, and the uniqueness of the keys (`r length(checks)` checks, all passing at render time). @tbl-columns describes the columns and what was done to each on import.

```{r}
#| label: tbl-columns
#| tbl-cap: "Columns of the data file, their role, and how each was treated on import."
tibble::tribble(
  ~Column, ~Role, ~`Cleaned type`, ~`Note on import`,
  "DepartamentoCodigo", "Identifier", "Character key", "Two-digit DANE code, leading zero kept",
  "Departamento", "Identifier", "Label", "Name; one department has two spellings",
  "MunicipioCodigo", "Identifier", "Character key", "Five-digit DANE code; `#TODOS` marks a department aggregate",
  "Municipio", "Identifier", "Label", "Name; not stable across years",
  "Año", "Identifier", "Integer", "Stored as text with a thousands separator (2,024)",
  "IRCA", "Measurement", "Numeric 0–100", "Annual index, whole unit",
  "Nivel de riesgo", "Measurement", "5-level factor", "Two spellings per band mapped to one",
  "IRCAurbano", "Measurement", "Numeric 0–100", "Urban area; `ND` read as missing",
  "Nivel de riesgo urbano", "Measurement", "5-level factor", "Missing when the value is `ND`",
  "IRCArural", "Measurement", "Numeric 0–100", "Rural area; `ND` read as missing",
  "Nivel de riesgo rural", "Measurement", "5-level factor", "Missing when the value is `ND`"
) |> kable()
```

Completeness needs a reference set of units that ought to be present. Two are used. The *internal* frame is every municipality code that appears at least once, crossed with every year that appears at least once; it asks whether the file is balanced on its own terms. The *external* frame is the municipal register of the DANE Marco Geoestadístico Nacional 2020 [@danemgn2020], which lists `r fmt_n(nrow(register))` units, minus the `r length(anm_codes)` non-municipalised areas of Amazonas, Guainía and Vaupés, which have no municipal administration and never appear in the file; that leaves `r fmt_n(n_register_muni)` municipalities and asks whether the country is covered. The register is a 2020 snapshot, so the few municipalities created after 2007 slightly overstate the external shortfall in early years. @fig-map shows the measurement and the coverage on that register.

TODO: make the figure caption shorter

```{r}
#| label: fig-map
#| fig-cap: !expr 'paste0("Left: annual IRCA by municipality in ", map_year, ", the year with the most municipalities reported, classed into the risk bands of Resolution 2115. Right: number of years, out of ", n_years, ", in which each municipality appears in the file; municipalities that never appear are outlined in orange. Non-municipalised areas are not expected in the file. The Archipelago of San Andrés lies outside the frame. Boundaries: DANE 2020.")'
#| fig-width: 6.5
#| fig-height: 4.6
by_year_ext <- frame_external |> group_by(year) |>
  summarise(missing = sum(!observed), observed = sum(observed), .groups = "drop")
stopifnot(all(by_year_ext$missing + by_year_ext$observed == n_register_muni))
map_year <- by_year_ext |> slice_max(observed, n = 1, with_ties = FALSE) |> pull(year)

mpio_sf <- st_read(gpkg_file, layer = "municipalities", quiet = TRUE)
dept_sf <- st_read(gpkg_file, layer = "departments", quiet = TRUE)
years_obs <- frame_external |> group_by(mpio_code) |> summarise(n_obs = sum(observed), .groups = "drop")

map_sf <- mpio_sf |>
  left_join(muni |> filter(year == map_year) |> select(mpio_code, irca), by = "mpio_code") |>
  left_join(years_obs, by = "mpio_code") |>
  mutate(anm = mpio_code %in% anm_codes,
         risk = cut(irca, risk_breaks, labels = risk_levels, right = TRUE),
         risk = as.character(risk),
         risk = case_when(anm ~ "Not expected", is.na(risk) ~ "No record", TRUE ~ risk),
         risk = factor(risk, levels = c(risk_levels, "No record", "Not expected")),
         n_obs = if_else(anm, NA_integer_, n_obs))
never_sf <- filter(map_sf, !anm, n_obs == 0)

base_map <- function() {
  list(geom_sf(data = dept_sf, fill = NA, colour = "#52514e", linewidth = 0.18),
       coord_sf(xlim = c(-79.2, -66.8), ylim = c(-4.3, 12.6), expand = FALSE),
       theme_void(base_size = 7.5),
       theme(legend.position = "bottom", legend.key.size = unit(0.3, "cm"),
             legend.text = element_text(size = 6.5), legend.title = element_text(size = 7),
             plot.margin = margin(0, 2, 0, 2)))
}
p_irca <- ggplot() +
  geom_sf(data = map_sf, aes(fill = risk), colour = "white", linewidth = 0.04) +
  scale_fill_manual(values = c(setNames(blue_ramp, risk_levels), "No record" = grey_none,
                               "Not expected" = grey_na),
                    name = paste("IRCA", map_year), drop = FALSE) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE, title.position = "top")) +
  base_map()
p_obs <- ggplot() +
  geom_sf(data = map_sf, aes(fill = n_obs), colour = "white", linewidth = 0.04) +
  geom_sf(data = never_sf, fill = NA, colour = col_rural, linewidth = 0.45) +
  scale_fill_gradient(low = blue_ramp[1], high = blue_ramp[5], limits = c(0, n_years),
                      breaks = c(0, 6, 12, 18), na.value = grey_na, name = "Years in file") +
  guides(fill = guide_colourbar(title.position = "top", barwidth = unit(3, "cm"),
                                barheight = unit(0.25, "cm"))) +
  base_map()
p_irca + p_obs
```

{{< pagebreak >}}

TODO: restrict the linewidth to 80 characters

# Question 1: What is the observational unit?

TODO: Give meaningful names to the sections instead of the questions I gave

## What a row looks like

```{r}
#| include: false
n_rows <- nrow(wq); n_cols <- ncol(raw)
ex_year <- year_max
ex_rows <- bind_rows(
  raw |> filter(MunicipioCodigo == "76001", Año %in% c("2,023", "2,024")),
  raw |> filter(DepartamentoCodigo == "76", MunicipioCodigo != "#TODOS",
                Año == "2,024", IRCArural == "ND") |> slice(1),
  raw |> filter(DepartamentoCodigo == "76", MunicipioCodigo == "#TODOS", Año == "2,024")
) |> select(DepartamentoCodigo, MunicipioCodigo, Municipio, Año, IRCA, IRCAurbano, IRCArural)
```

The file has `r fmt_n(n_rows)` rows and `r n_cols` columns. @tbl-example reproduces four rows verbatim. Each row names a place and a year and reports three indices: one for the whole territorial unit, one for its urban area and one for its rural area, each accompanied in the full file by a risk band derived from it. The same municipality code recurs year after year, so a row is not a place but a place in a year. The name attached to a code can change between years while the code stays fixed. The token `ND` stands in for a component that was not reported. The fourth row carries `#TODOS` in place of a municipality: it describes the department as a whole for that year, and `r fmt_n(nrow(dept))` such rows are interleaved with the `r fmt_n(nrow(muni))` municipal rows.

```{r}
#| label: tbl-example
#| tbl-cap: "Four rows of the file, reproduced verbatim: the same municipality in two consecutive years under two names, a municipality with no rural index, and the department-level aggregate for the same department and year."
kable(ex_rows)
```

## What identifies a row

```{r}
#| include: false
key_check <- function(df, cols, label) {
  k <- df[cols]
  tibble(`Candidate key` = label,
         `Distinct values` = fmt_n(nrow(distinct(k))),
         Rows = fmt_n(nrow(k)),
         `Duplicated rows` = fmt_n(sum(duplicated(k))))
}
raw_muni <- raw |> filter(MunicipioCodigo != "#TODOS")
raw_dept <- raw |> filter(MunicipioCodigo == "#TODOS")
keys_tbl <- bind_rows(
  key_check(raw_muni, "MunicipioCodigo", "Municipality code"),
  key_check(raw_muni, c("Municipio", "Año"), "Municipality name + year"),
  key_check(raw_muni, c("Departamento", "Municipio", "Año"), "Department name + municipality name + year"),
  key_check(raw_muni, c("MunicipioCodigo", "Año"), "Municipality code + year"),
  key_check(raw_dept, c("DepartamentoCodigo", "Año"), "Department code + year (aggregate rows)")
)
n_dup_name_year <- sum(duplicated(raw_muni[c("Municipio", "Año")]))
```

@tbl-keys tests candidate identifiers on the municipal rows. A municipality code alone recurs across years. A municipality name with a year is not unique either, because `r n_names_multi_code` names are shared by municipalities in different departments. Adding the department name makes the key unique, but names are not stable: municipality 76001 appears as "Santiago de Cali" and as "Cali", and over the period more than one name is attached to `r n_codes_multi_name` of the `r fmt_n(n_muni_codes)` codes. The municipality code with the year identifies every municipal row exactly once, and because the first two digits of every municipality code equal the department code, the department adds no information. On the aggregate rows, department code with year is likewise unique.

```{r}
#| label: tbl-keys
#| tbl-cap: "Candidate identifiers tested on the municipal rows and, in the last line, on the department aggregate rows."
kable(keys_tbl, align = "lrrr")
```

The municipal rows cover `r fmt_n(n_muni_codes)` municipality codes in `r n_dept_codes` departments over `r n_years` years, `r year_min` to `r year_max`.

## Answer

TODO: Give the document a more flowing style instead of question and answer

The observational unit is the **municipality-year**. One row records, for one municipality in one calendar year, the annual IRCA of the municipality as a whole and of its urban and rural areas, with the risk band of each. The department-year rows marked `#TODOS` are a second, coarser set of units derived from the same surveillance; they must be separated from the municipal rows before any count or average, or departments will be counted twice.

It is equally important what the unit is not. The IRCA of a municipality-year is a consolidation of many water samples taken over the year, but the samples, the supply systems, the sampling points and the months are not in the file, and the number of samples behind a value is unknown. A row is therefore the annual summary of a municipality's surveillance, not an observation of any water supply. For any join or panel construction the numeric code, not the name, is the key.

{{< pagebreak >}}

# Question 2: How complete are the data?

Incompleteness can occur at three levels, and the answer differs at each. A row can exist but lack a value (a missing value). A municipality can lack a row in a year in which other municipalities have one (a missing unit). A year can be absent altogether, or absent from the series of a particular municipality (a missing year).

## Missing values within rows

```{r}
#| include: false
n_empty <- 0L  # asserted in prepare.R
nd_pat <- muni |>
  mutate(u = is.na(irca_urb), r = is.na(irca_rur)) |>
  count(u, r) |> mutate(share = n / sum(n))
pct_rural_nd <- mean(is.na(muni$irca_rur)); pct_urban_nd <- mean(is.na(muni$irca_urb))
pct_all_three <- nd_pat |> filter(!u, !r) |> pull(share)
n_both_nd <- nd_pat |> filter(u, r) |> pull(n)
pct_dept_rural_nd <- mean(is.na(dept$irca_rur))
nd_year <- muni |> group_by(year) |>
  summarise(Urban = mean(is.na(irca_urb)), Rural = mean(is.na(irca_rur)),
            Both = mean(is.na(irca_urb) & is.na(irca_rur)), .groups = "drop") |>
  pivot_longer(-year, names_to = "component", values_to = "share") |>
  mutate(component = factor(component, levels = c("Rural", "Urban", "Both")))
urban_peak <- nd_year |> filter(component == "Urban") |> slice_max(share, n = 1)
rural_first <- nd_year |> filter(component == "Rural", year == year_min) |> pull(share)
rural_pat <- muni |> group_by(mpio_code) |> summarise(p = mean(is.na(irca_rur)), .groups = "drop")
n_rural_always <- sum(rural_pat$p == 1); n_rural_never <- sum(rural_pat$p == 0)
n_rural_inter <- sum(rural_pat$p > 0 & rural_pat$p < 1)
rn <- muni |> filter(is.na(irca_rur), !is.na(irca_urb)) |> mutate(d = abs(irca - irca_urb))
pct_total_eq_urban <- mean(rn$d < 0.051); n_total_ne_urban <- sum(rn$d > 1)
```

TODO: If n_empty comes prom another script load it instead of just defining it

No cell is empty (`r n_empty` empty cells), and the overall IRCA and its band are never missing (`r n_irca_nd` rows). Missingness lives entirely in the urban and rural components, encoded by `ND`, always for the value and its band together. In the municipal rows the rural index is missing in `r fmt_pct(pct_rural_nd)` of rows and the urban index in `r fmt_pct(pct_urban_nd)`; only `r fmt_pct(pct_all_three)` of rows carry all three indices, and `r fmt_n(n_both_nd)` rows carry an overall index with neither component. Department aggregates are better filled: the rural index is missing in `r fmt_pct(pct_dept_rural_nd)` of them. Every reported band agrees with the band implied by its value (`r label_agreement$n_disagree` disagreements in `r fmt_n(label_agreement$n)` values).

The rural gap has changed over time (@fig-nd-year). It affected `r fmt_pct(rural_first)` of rows in `r year_min`, fell to around 40% by 2010, and has stayed above half since 2016. The urban index is rarely missing except in 2016 and 2017, peaking at `r fmt_pct(urban_peak$share)` in `r urban_peak$year`, when both components go missing together, which suggests a reporting artefact rather than a change in sampling. The rural gap is intermittent rather than structural: `r n_rural_always` municipalities never report a rural index, `r n_rural_never` always do, and `r fmt_n(n_rural_inter)` alternate.

Two things distinguish `ND` from a missing unit. Definitionally, a row with `ND` still carries an overall index, whereas an absent municipality-year carries nothing. Empirically, `ND` does not always mean that no rural samples were taken: among the `r fmt_n(nrow(rn))` rows with a rural `ND` and an urban value, the overall index equals the urban index in `r fmt_pct(pct_total_eq_urban)` of cases, but in `r n_total_ne_urban` rows the two differ by more than one point, so rural samples entered the overall index even though the rural component is reported as unavailable.

```{r}
#| label: fig-nd-year
#| fig-cap: "Share of municipal rows in which the urban index, the rural index, or both are reported as ND, by year."
#| fig-width: 6.5
#| fig-height: 2.4
nd_cols <- c(Rural = col_rural, Urban = col_urban, Both = col_both)
ggplot(nd_year, aes(year, share, colour = component, linetype = component)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  scale_colour_manual(values = nd_cols, name = NULL) +
  scale_linetype_manual(values = c(Rural = "solid", Urban = "solid", Both = "22"), name = NULL) +
  scale_x_continuous(breaks = years, expand = c(0.02, 0)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.8),
                     expand = c(0, 0)) +
  labs(x = NULL, y = "Share of rows with ND") +
  theme(legend.position = "bottom", legend.margin = margin(0, 0, 0, 0),
        axis.text.x = element_text(size = 6.5))
```

## Missing observational units

```{r}
#| include: false
frames_tbl <- tibble(
  Units = c("Municipality-years, internal frame", "Municipality-years, external frame",
            "Municipalities, external frame", "Department-years"),
  expected = c(nrow(frame_internal), nrow(frame_external), n_register_muni, n_dept_codes * n_years),
  observed = c(sum(frame_internal$observed), sum(frame_external$observed), n_muni_codes, nrow(dept))
) |> mutate(missing = expected - observed, share = missing / expected)
stopifnot(all(frames_tbl$expected == frames_tbl$observed + frames_tbl$missing))
pct_missing_int <- frames_tbl$share[1]; pct_missing_ext <- frames_tbl$share[2]
n_missing_ext <- frames_tbl$missing[2]
best_year <- by_year_ext |> slice_min(missing, n = 1, with_ties = FALSE)
worst_year <- by_year_ext |> slice_max(missing, n = 1, with_ties = FALSE)
by_dept_ext <- frame_external |> group_by(dept_code) |>
  summarise(n_m = n_distinct(mpio_code), missing = sum(!observed), share = mean(!observed),
            .groups = "drop") |>
  left_join(dept_short, by = "dept_code") |> arrange(desc(share))
n_dept_complete <- sum(by_dept_ext$missing == 0)
choco <- filter(by_dept_ext, dept_code == "27")
choco_first <- min(wq$year[wq$dept_code == "27"])
never_tbl <- register_muni |> filter(!mpio_code %in% muni$mpio_code) |>
  left_join(dept_short, by = "dept_code") |> arrange(dept_label, mpio_name)
never_txt <- never_tbl |> group_by(dept_label) |>
  summarise(txt = paste0(dept_label[1], ": ", paste(mpio_name, collapse = ", ")), .groups = "drop") |>
  pull(txt) |> paste(collapse = "; ")
n_never <- nrow(never_tbl)
my_dy <- distinct(muni, dept_code, year); dy <- distinct(dept, dept_code, year)
n_dept_inconsistent <- nrow(anti_join(my_dy, dy, by = c("dept_code", "year"))) +
  nrow(anti_join(dy, my_dy, by = c("dept_code", "year")))
n_missing_dy <- frames_tbl$missing[4]
```

@tbl-frames counts the units that are expected and observed under each frame. Against the file's own frame, `r fmt_pct(pct_missing_int, 0.1)` of municipality-years are absent; against the national register, `r fmt_pct(pct_missing_ext, 0.1)`, because `r n_never` municipalities never appear at all (`r never_txt`). Coverage improved from `r year_min` to `r best_year$year`, when only `r best_year$missing` municipalities were absent, and has weakened since 2020; `r worst_year$year` is the least complete year, with `r worst_year$missing` absent. Department aggregates are absent in `r n_missing_dy` department-years, exactly those in which the department reported no municipality at all (`r n_dept_inconsistent` exceptions), so the aggregate rows add no coverage of their own.

```{r}
#| label: tbl-frames
#| tbl-cap: "Expected and observed units under the internal frame (codes present at least once by years present at least once) and the external frame (DANE 2020 municipal register, non-municipalised areas excluded)."
frames_tbl |>
  transmute(Units, Expected = fmt_n(expected), Observed = fmt_n(observed),
            Missing = fmt_n(missing), `Share missing` = fmt_pct(share, 0.1)) |>
  kable(align = "lrrrr")
```

The gaps are concentrated (@fig-heatmap). Only `r n_dept_complete` departments are complete over the whole period. Chocó has no rows of any kind before `r choco_first` and is missing `r fmt_pct(choco$share)` of its municipality-years, `r fmt_n(choco$missing)` of the `r fmt_n(n_missing_ext)` absent nationally; Guainía, Vaupés, Guaviare, Córdoba and Amazonas each lack more than a fifth.

```{r}
#| label: fig-heatmap
#| fig-cap: "Share of each department's municipalities absent from the file in each year, on the external frame. Departments are ordered by their overall share missing; the number of municipalities in the register follows each name."
#| fig-width: 6.5
#| fig-height: 3.6
dept_year_ext <- frame_external |> group_by(dept_code, year) |>
  summarise(share_missing = mean(!observed), .groups = "drop") |>
  left_join(by_dept_ext |> select(dept_code, dept_label, n_m, share), by = "dept_code") |>
  mutate(lab = paste0(dept_label, " (", n_m, ")"))
lab_order <- dept_year_ext |> distinct(lab, share) |> arrange(share) |> pull(lab)
ggplot(dept_year_ext, aes(year, factor(lab, levels = lab_order), fill = share_missing)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "#f4f3ef", high = blue_ramp[5], limits = c(0, 1),
                      labels = percent_format(accuracy = 1), name = "Absent") +
  scale_x_continuous(breaks = years, expand = c(0, 0)) +
  labs(x = NULL, y = NULL) +
  theme(panel.grid = element_blank(), axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6, angle = 45, hjust = 1),
        legend.key.height = unit(0.5, "cm"), legend.key.width = unit(0.25, "cm"))
```

## Missing years and panel balance

```{r}
#| include: false
n_calendar_gaps <- length(setdiff(seq(year_min, year_max), years))
series <- frame_internal |> group_by(mpio_code) |>
  summarise(n = sum(observed), first = min(year[observed]), last = max(year[observed]),
            .groups = "drop") |>
  mutate(interior = (last - first + 1) - n,
         late = first > year_min, early = last < year_max, gaps = interior > 0,
         pattern = case_when(n == n_years ~ "Complete series",
                             late & !early & !gaps ~ "Starts late only",
                             !late & early & !gaps ~ "Ends early only",
                             !late & !early & gaps ~ "Interior gaps only",
                             TRUE ~ "Two or more of the above"))
pattern_levels <- c("Complete series", "Starts late only", "Ends early only",
                    "Interior gaps only", "Two or more of the above")
series_tbl <- series |>
  mutate(pattern = factor(pattern, levels = pattern_levels)) |>
  group_by(pattern, .drop = FALSE) |>
  summarise(n_muni = n(), missing = sum(n_years - n), .groups = "drop") |>
  mutate(share = n_muni / sum(n_muni))
stopifnot(sum(series_tbl$n_muni) == n_muni_codes,
          sum(series_tbl$missing) == sum(!frame_internal$observed))
n_full <- series_tbl$n_muni[1]; pct_full <- series_tbl$share[1]
n_late <- sum(series$late); n_early <- sum(series$early); n_gaps <- sum(series$gaps)
n_gap_years <- sum(series$interior)
balanced <- sapply(c(2010, 2015, 2019), function(s) {
  frame_external |> filter(year >= s) |> group_by(mpio_code) |>
    summarise(all = all(observed), .groups = "drop") |> pull(all) |> sum()
})
```

At the national level no year is missing: every calendar year from `r year_min` to `r year_max` is present (`r n_calendar_gaps` gaps). The file was downloaded in September 2026, so the absence of 2025 most likely reflects publication lag rather than a gap. At the municipal level the picture is different (@tbl-series). `r fmt_n(n_full)` municipalities (`r fmt_pct(pct_full)`) are observed in every year. Of the rest, `r n_late` enter after `r year_min`, `r n_early` drop out before `r year_max`, and `r n_gaps` have at least one interruption inside their series, which together account for `r fmt_n(n_gap_years)` interior gap-years. For anyone who needs a balanced panel, shortening the window helps only moderately: `r fmt_n(balanced[1])` of the `r fmt_n(n_register_muni)` register municipalities are observed in every year from 2010, `r fmt_n(balanced[2])` from 2015 and `r fmt_n(balanced[3])` from 2019.

```{r}
#| label: tbl-series
#| tbl-cap: "Continuity of each municipality's series on the internal frame. A series starts late if its first year is after 2007, ends early if its last year is before 2024, and has interior gaps if years are missing between its first and last."
series_tbl |>
  transmute(Pattern = as.character(pattern), Municipalities = fmt_n(n_muni),
            Share = fmt_pct(share), `Missing municipality-years` = fmt_n(missing)) |>
  kable(align = "lrrr")
```

## Answer

Are there missing values? Not in the overall index, which is present in every row, but yes in the components: the rural index is unavailable in about half of the municipal rows, and `ND` sometimes hides rural samples that did enter the overall index. Are there missing units? Yes: about 6% of the municipality-years implied by the national register are absent, `r n_never` municipalities never appear, and the shortfall is concentrated in Chocó and the Amazon and Orinoco departments. Are there missing years? Not at the national level, where 2007 to 2024 is continuous, but a third of municipalities have a series that starts late, ends early or is interrupted. Analyses of the overall index at the national level can proceed with few reservations; analyses of the urban or rural components, comparisons across departments, and municipal time series must treat the missingness explicitly.

```{r}
#| label: check-abstract
#| include: false
# The abstract states rounded figures in words; make sure they still hold.
stopifnot(
  abs(pct_rural_nd - 0.5) < 0.05,          # "about half"
  abs(pct_missing_ext - 0.06) < 0.01,      # "about 6%"
  abs(pct_full - 2 / 3) < 0.04,            # "two thirds"
  choco$share > 0.8,                       # "more than four fifths"
  n_irca_nd == 0, n_calendar_gaps == 0
)
```

# References {.unnumbered}

TODO: limit the number of references to 5

::: {#refs}
:::
